Towards an Ecological Trait-data Standard

### 2018-10-11 manuscript/manuscript.Rmd
revise manuscript: update vocabulary description, discussion, bibliography

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2] \footnotemark[3]
- name: Malte Jochum \footnotemark[4]
- name: Gaëtane Le Provost \footnotemark[3]
- name: Andreas Ostrowski \footnotemark[6]
- name: Caterina Penone \footnotemark[4]
- name: David Fichtmüller \footnotemark[5]
- name: Martin M. Gossner \footnotemark[8]
- name: Anton Güntsch \footnotemark[5]
- name: Birgitta König-Ries\footnotemark[6]
- name: Pete Manning \footnotemark[3]
- name: Nadja K. Simons \footnotemark[7]
csl: amnat.csl
fontsize: 11pt
geometry: margin=1.4in
output:
  word_document:
    fig_caption: yes
    fig_width: 6
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
bibliography: div05.bib
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[2]{Institute of Linguistics and Literary Studies, Technische Universität Darmstadt, Darmstadt, Germany}
\footnotetext[3]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
\footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}

@@

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis17; @mantle12; @strobel18] and citizen science initiatives (e.g. www.markmybird.org), 
For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).

@@

Whiel trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.  
To make sense of trait data in the context of more general databases, a consensus definition of trait data is therefore necessary.

Trait data have been defined by @garnier17 as following an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e. trait values) describing the physical phenotypic characteristics relating to fitness and performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or its parts) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
The entity at which the trait was assessed may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

@@

Initiatives for standardisation (e.g. ontologies and data standards) and platforms for data management (e.g. database and data management platforms) provide great visibility, accessability and improved interoperability of datasets, but raise relatively high thresholds for data and metadata preparation. 
This contrasts with the growing numer of openly accessible trait datasets in public file-servers, which hardly any of the aforementioned trait datasets or databases for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards. As it stands, re-using these data in larger compilations is error-prone and inefficient and the potential for a broad synthesis is diminished.
In sum, we attest a gap in the trait-data structures developed for data curators and managers and the unstructured data input by data providers.

One likely reason for this is complexity: the proposed data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice. In the eyes of the data-provider, in most cases, ancillary co-factors can be appended as extra columns to the dataset.
<!--If the traits are assembled for a specific research question with just few ancillary co-factors, it may be safer to append these as columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road. This way, if datasets of different sources are merged, the related information is readily available without the risk of breaking the reference to an external datasheet. -->
The other reason is lack of awareness of the need for trait-data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management. In addition, complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.

@@

Another solution for data-users to access trait data in a structured way is offered by decentralised tools and toolchains. For example,
- r packages
FENNEC [@ankenbrand18], an online tool or self-hosted service that can extract trait information from multiple sources for a target species community.

@@

Despite centralised database efforts, no unified definition of traits as a data type with a well-defined exchange format or data structure has emerged across the multiple sub-disciplines of trait-based research.

However, unlike for other data types like for instance genetic data or movement data (@@), no unified format for trait data has successfully been established for application in all use-cases.

Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards.
A more widespread implementation of ontologies would advance the possibilities to aggregate datasets into databases and reduce noise and uncertainty. 
To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
For example, the accessibility of ontologies will increase if open *Application Programming Interfaces (APIs)* are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. 
Software then can assist researchers in linking own data to globally defined concepts.

Therefore we must better integrate these distributed data into the global biodiversity data-management life-cycle by making data providers more aware of data visibility, accessibility, interoperability and re-useability. By establishing this as a best-practice standard in trait-based research, trait-data will eventually fulfil the FAIR guiding principles for scientific data [@wilkinson16].

In the following chapter, we propose the development of vocabularies that lower the thresholds and offer high pay-off in the visibility and reuse of published data.

@@

As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research. 
The standard vocabulary is accessible at https://ecologicaltraitdata.github.io/ETS/.
The aim of the vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement detail. 
Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in three major use cases:

1. in  two-dimensional datasets, e.g. for labelling project-specific data, for local use and exchange with collaborators, or for publication of standardised primary data on open-access data repositories, 
2. as a consensus vocabulary when compiling data from distributed sources into databases, e.g. to map standardised columns and refer to taxa and trait definitions in a uniform way, and
3. as an exchange format of data between web services and software tools, e.g. when providing database queries from a webservice or defining input and output formats of software packages.

By using this standard vocabulary, authors can ensure that the description of trait measurements that are uploaded to distributed data repositories will be unambiguous and generally applicable. It will facilitate re-use of data for future data aggregation initiatives and data synthesis and ensure long-term accessibility. By adopting the standard vocabulary, trait-data will fulfill the FAIR principles for scientific data [@wilkinson16]. 
We describe best practice principles and give examples for the application of the vocabulary in Supplementary Material B.

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).
We paid particular consideration to existing data standards (particularly @kattge11; @kattge11a; @garnier17; @parr16) to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. 
<!--Besides being used for the publication of datasets, the standard vocabulary could be imposed in webservices or download tools, e.g. APIs that provide direct access to online databases. -->
<!--
## Why use the vocabulary

@@

The core terms describe a dataset according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e. quality) observed on an individual or population (i.e. entity), of a biological taxon.

We are expanding on this definition: quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig. 1*a*), by aggregating replicated measurements on multiple entities (Fig. 1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g. databases, Fig. 1*c*). This encompasses a wide range of data types, including numeric, integer, intervals or ratios, or even values from an ordinal scale. Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
\begin{figure}[t]
\centering
\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
\caption{Types of ecological trait data assume different entities or reported quantities. a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g. reproductive rates, life spans); b) aggregated traits are reported as means taken on multiple measures of members of a taxon; c) quantities may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g. a sex) as the entity of description; d) Qualitative traits are categorical or binary descriptors of the entire species or higher taxonomic group.}

@@

<!--Beyond these core observations, further information might be available that are related to the taxonomic assignment, or that put the reported fact, measurement or sampling event in a broader observation context (including geolocation and date information). 
These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data. For this case, -->
We offer three extensions of the core vocabulary ("Taxon", "Measurement or Fact", and "Occurrence"), which may simply be added as extra columns to the core dataset (see sections below for extensions and best practice when publishing primary data).
Additional terms are provided for metadata and for relating trait names to definitions and external ontologies or thesauri (see section on metadata below).

Some of the terms expand and refine terms of the Darwin Core and its Extensions (http://rs.tdwg.org/dwc/terms/) or the Glossary of EOL (http://eol.org/info/516). Wherever our vocabular refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively.
Any further columns on co-variates or measurement context, or other dataset-specific content, might draw from terms provided by Darwin Core and it's extensions, or be added as user-defined columns. However, these should be clearly specified in the meta-information accompanying the dataset. 
The scope of the vocabulary may not yet cover all aspects of morphological and evolutionary perspectives. 
Also, information about interactions between species are not within the scope of the Entity-Quality Model, but may easily be combined with trait data by using other extensions of DwC. 
The vocabulary proposed is intended to form the base of a standard exchange format.
Therefore, we invite researchers to contribute to the next iterations of the standard vocabulary and develop own applications and ontologies that interact with it.

@@

To qualify as trait data according to the definition provided above, where each row is the reported measurement or fact for a single observation, the following columns are required at minimum  (Fig. 2 *b*):
1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
2. a descriptive trait name (`traitName`) that links to a well-defined definition; 
3. the scientific taxon name for which the measurement or fact was obtained (`scientificName`). 
For these core values, unambiguous and self-explanatory vocabularies for trait names and taxa are recommended. 
However, to ensure compatibility with existing databases or analytical code, it might be necessary to use abbreviations or user-specific identifiers for `scientificName`  and `traitName` instead. 
<!--### Relate to global taxon and trait definitions -->
In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up table of traits. 
This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively). 
For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass'.  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practised in TRY; @kattge11]. This redundancy of data allows for continuity for data owners while also ensuring quality checks and comparability for the data user.

@@

Beyond measurement units or higher taxon information, further information might be available that may not be core data, but are related to the individual or specimen, or to the reported fact, measurement or sampling event. 
The data standard provides three extensions of the vocabulary that should be used to describe this information \mbox{(Fig. 2*d*)}:

- The `Taxon` extension provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.  (http://ecologicaltraitdata.github.io/ETS/#extension-taxon)
- The `MeasurementOrFact` extension  provides terms to describe information at the level of single measurements or reported facts, such as the original literature from where the value is cited, the method of measurement or statistical method of aggregation. It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g. variation in measurement method) or aggregated values (e.g. statistical method applied), as well as the source of reported facts (e.g. literature source or expert reference). (https://ecologicaltraitdata.github.io/ETS/#extension-measurement-or-fact)
- The `Occurrence` extension contains vocabulary to describe information on the level of individual specimens, such as sex, life stage or age. This also includes the method of sampling and preservation, as well as date and geographical location, which provides an important resource to analyse trait variation due to differences in space and time. (https://ecologicaltraitdata.github.io/ETS/#extension-occurrence)

<!--Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
 If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).  
Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level. This information can then be used for trait inference at higher taxonomic resolution. <!-- using hierarchical probabilistic matrix factorisation [@schrodt15]. -->
These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets, thus avoiding redundancy and duplication of content. A unique identifier would link to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`). 
Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g. a GBIF URI or a museum collection code references the precise specimen from which the measurement was taken [@groom17; @guntsch17].

@@

Wherever possible, the column `traitID` should point to a publicly available, unambiguous trait definition in a published ontology (using its URI). 
If no globally available trait definition exists as an external reference, trait datasets should always be accompanied by a dataset-specific list of traits as part of the metadata or as an accompanying data table.
Such a controlled vocabulary would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format of measured values or reported facts (e.g. units or legit factor levels). Ideally, this definition refers to or refines terms from published trait ontologies.
By providing a minimal vocabulary for trait lists (see https://ecologicaltraitdata.github.io/ETS/#terms-for-trait-definitions), we hope to facilitate the unambiguous definition of traits for trait datasets.
This vocabulary might also prove useful for the future publication of trait ontologies. 
<!--See our online reference for recommendations how to prepare a traitlist for  (https://ecologicaltraitdata.github.io/ETS/thesauri.html). We also provide an example thesaurus on Github (http://ecologicaltraitdata.github.io/TraitDataList/), which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project [@fischer10]. -->

Information about the authorship and ownership of the data and the terms of use should be considered when sharing and working with trait datasets.
We define a vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) that allows trait data to be related to authors and owners, while also stating a bibliographic reference and license model. 
In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata [e.g. in a separate metadata file, possibly applying the ecological metadata language, EML, @knb11]. 
In cases where individual data from different sources are compiled into a trait database, these information must be provided at the measurement level to maintain the reference to the original data provider and conditions of re-use of these data. This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.

<!--Open access to science data from publicly funded research is a declared goal of international and national research organisations [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12] and increasingly required by journals and funding agencies to warrant reproducibility of science and foster public benefit. 
Also, factual data such as trait measurements, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@carroll15; @creativecommons].
However, there is still no broad consensus achieved in the scientific community whether data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship. -->
<!-- cP: Maybe we should put in the appendix an example of dataset+metadata? This should probably go in the metadata section 
FDS: to consider!! -->

## Best practice for primary data publication

We focus this section on the first of the three use-cases for the ecological trait-data standard vocabulary: The publication of project-specific primary data.  We assume that data managers and software developers who apply the vocabulary in use-case 2 and 3 are more fluent with the use of exchange formats and standard definitions and have established own best-practice principles.

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
- name: David Fichtmüller \footnotemark[5]
- name: Martin M. Gossner \footnotemark[8]
- name: Anton Güntsch \footnotemark[5]
- name: Malte Jochum \footnotemark[4]
- name: Birgitta König-Ries\footnotemark[6]
- name: Gaëtane Le Provost \footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[6]
- name: Caterina Penone \footnotemark[4]
- name: Nadja K. Simons \footnotemark[7]
csl: amnat.csl
fontsize: 11pt
geometry: margin=1.4in
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

\footnotetext[1]{Corresponding author: florian.dirk.schneider@gmail.com}
\footnotetext[3]{Senckenberg Biodiversity and Climate Research Centre (BiK-F) Frankfurt am Main, Germany}
\footnotetext[5]{Botanic Garden and Botanical Museum Berlin, Freie Universität Berlin, Berlin, Germany}
\footnotetext[8]{Forest Entomology, Swiss Federal Research Institute WSL, Birmensdorf, Switzerland} <!--  Zürcherstrasse 111, CH-8903 -->
\footnotetext[4]{Institute of Plant Sciences, University of Bern, Bern, Switzerland}
\footnotetext[6]{Department of Mathematics and Computer Science, Friedrich-Schiller-Universität Jena, Jena, Germany}
\footnotetext[7]{Department of Ecology and Ecosystem Management, Technische Universität München, Freising, Germany}

@@

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g. www.markmybird.org), 
For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).

@@

While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.

To make sense of trait data in the context of more general databases, a consensus definition of trait data has been provided by  @garnier17: 
The define trait data to follow an entity-quality model (EQ), where a trait is defined as 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative *measurements* or qualitative *facts* (i.e. trait values) describing the physical phenotypic characteristics relating to fitness and performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or its parts) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
The entity at which the trait was assessed may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

@@

In sum, we attest a gap in the trait-data structures developed for data curators and managers and the data input by data providers. Hardly any of the aforementioned trait datasets or databases for birds, amphibians, mammals or invertebrates employs the described standard terminologies, ontologies or semantic web standards. As it stands, re-using these data in larger compilations is error-prone and inefficient and the potential for a broad synthesis is diminished.

One likely reason for this is complexity: the proposed data structures are designed for multi-layered, relational databases rather than for standalone datasets for which a two-dimensional data table may suffice. In the eyes of the data-provider, in most cases, ancillary co-factors can be appended as extra columns to the dataset.
The other reason is lack of awareness of the need for trait-data standardisation among data providers: many providers are not trained in the demands of biodiversity data-management. In addition, complying with what may be non-intuitive data structures is an investment without clear incentive or immediate pay-off, and hardly affordable for small and intermediate-size research projects.

@@

Another solution for data-users to access trait data in a structured way is offered by decentralised tools and toolchains to facilitate the use and analysis of trait-data. 
For instance the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB. The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species$\times$traits matrix. 
The package 'NatDB' [@pearse18] collates distributed trait data and harmonizes them into a unified format, and assists in browsing and filtering these data. 
FENNEC [@ankenbrand18], an online tool or self-hosted service that can extract trait information from multiple sources for a target species community.

@@

Interlinkage and accessibility of ontologies can be much improved to fulfil semantic web standards. 
A more widespread implementation of ontologies would advance the possibilities to aggregate datasets into databases and reduce noise and uncertainty. 
To achieve this, the use of ontologies and thesauri must be incentivised and facilitated for individual researchers.
For example, the accessibility of ontologies will increase if open APIs are provided as a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. 
Software then can assist researchers in linking own data to globally defined concepts.

To bring all these tools for datamanagement together, a common vocabulary and definition of consensus concepts and exchance formats is required that connects to all steps of the data life-cycle.
Despite centralised data management efforts, however, no unified definition of traits as a data type with a well-defined exchange format or data structure has emerged across the multiple sub-disciplines of trait-based research. 
In the following chapter, we propose the a unified vocabulary for trait-data that can form the basis of such an exchange format. The simplicity of this standard terminology will lower the thresholds and offer high pay-off in the visibility and reuse of published data. 
By establishing this as a best-practice standard in trait-based research, trait-data will eventually fulfil the FAIR guiding principles for scientific data [@wilkinson16].

@@

As a response to the challenges outlined above, we propose a versatile vocabulary for trait-based ecological research. 
The standard vocabulary is accessible at https://terminologies.gfbio.org/terms/ets/pages/.
The aim of this vocabulary is to cover the variety of trait-based approaches and their different degrees of measurement details. 
Rather than describing a data structure for relational databases, the vocabulary is intended as a more inclusive terminology, that can be used in three major use cases: 
1. by data providers: for publication of standardised primary data on open-access data repositories, or for labelling project-specific data for local use and exchange with collaborators, e.g. in two-dimensional data tables or project databases
2. by data users and curators: as a consensus vocabulary when compiling data from distributed sources into databases, e.g. to map standardised columns and refer to taxa and trait definitions in a uniform way, and
3. by data managers: as a data exchange format of between databases, web services and software tools, e.g. when providing database queries via a web service or defining input and output formats of software packages.

By using this standard vocabulary, authors can ensure that the description of trait measurements uploaded into distributed data repositories will be unambiguous and generally applicable. It will facilitate interoperability of published data and enable their re-use for future data aggregation initiatives, data synthesis and ensure long-term accessibility. By adopting the standard vocabulary, trait-data will support the FAIR principles for scientific data [@wilkinson16]. 
We describe best practice principles and give examples for the application of the vocabulary in Supplementary Material B.

In designing this vocabulary, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users and curators), and biodiversity informatics researchers (data managers).
We paid particular attention to existing data standards [particularly @madin07; @kattge11; @kattge11a; @garnier17; @parr16] to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management.

<!--
## Why use the vocabulary

@@

A trait dataset typically contains data according to the Entity-Quality model described above [@garnier17]: each entry describes a trait value (i.e. quality) observerd on an individual or population (i.e. entity), of a biological taxon. 
The trait values may be quantitative measurements or qualitative facts. 
Quantitative measurements are values obtained either by direct morphological, physiological or behavioural observations on single specimens (Fig. 1*a*), by aggregating replicated measurements on multiple entities (Fig. 1*b*) or by estimating the means or ranges for the respective taxon as reported in the literature or other published sources (e.g. databases, Fig. 1*c*). 
This encompasses a wide range of data types, including numeric, integer, intervals or ratios, or even values from an ordinal scale. Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
Beyond these core observations, further information might be available that further specify the taxon concept applied, provide detail on the measurement method, or that place the reported measurement in a broader observation context (including geolocation and date information).
These information may be useful for future analysis of the causal reasons of trait variation and should always be published along with the core data.

\begin{figure}[t]
\centering
\includegraphics[width=0.85\textwidth, angle=0]{figures/fig1.png}
\caption{Types of ecological trait data assume different entities or reported qualities: a) morphometric or morphological measurements of individual body features (lengths, areas, volumes, weights) or other quantities related to life history (e.g. reproductive rates, life spans); b) aggregated trait values are reported as means taken on multiple measures of members of a taxon; c) quantitative traits may be extracted from literature or existing databases, referring to the entire taxon (or a subset, e.g. a sex) as the entity of description; d) qualitative traits are categorical, ordinal or binary descriptors of the entire species or higher taxonomic level (also called 'facts').}

@@

Our vocabulary thus provides core terms to reflect the minimal Entity-Quality structure of trait data. 
We further offer three extensions to contain ancillary information on different layers of the core data ("Measurement or Fact", and "Occurrence", "Taxon"; see section on extensions below). 
All terms may be applied to describe columns of a data table, or to label content of a database (see Box 1 / Supplementary Material B for best practice principles and examples for publishing primary data).
Additional terms are provided for metadata and for providing trait concepts (see section on metadata below).  
Some of the terms expand and refine terms of DwC and its extensions (http://rs.tdwg.org/dwc/terms/) or the Glossary of EOL (http://eol.org/info/516). Wherever our vocabular refines or duplicates existing terms, we indicate this in the fields 'refines' or 'identical', respectively. The vocabulary can be mapped onto other existing ontologies (e.g. @madin07) or database structures (e.g. @kattge11a).
Aspects not covered by the terminology may draw from terms provided by DwC and it's extensions or be added as user-defined columns (which should be clearly specified in the meta-information accompanying the dataset).

@@

For trait data according to the Entity-Quality model, the following content is required at minimum (Fig. 2 *a&b*):
1. a value (column `traitValue`) and -- for numeric values -- a standard unit (`traitUnit`); 
2. a descriptive trait name (`traitName`) that links to a standardised definition (i.e. a concept); 
3. the scientific taxon name for which the measurement or fact was obtained (`scientificName`) that links to an accepted taxon concept.

<!--### Relate to global taxon and trait definitions -->
The trait name and taxon names as would ideally use unambiguous terms assigning both to clearly-defined concepts. 
However, to ensure compatibility with existing databases or analytical code, it might be necessary to use abbreviations or user-specific identifiers for `scientificName`  and `traitName` instead. 
In this case, it is essential to relate the user-defined names to a consensus standard of taxon names as well as a look-up table of traits. 
This is achieved by adding globally valid Uniform Resource Identifiers (URIs) for taxon (`taxonID`) and trait definitions (`traitID`), complemented by the human-readable verbatim accepted names (`ScientificNameStd` and `traitNameStd`, respectively, Fig. 2 *c*). 
For example, referring to GBIF Backbone Terminology, for *Bellis perennis*, the `taxonID` would be  `https://www.gbif.org/species/3117424`; the `traitID` for 'fruit mass' according to TOP Thesaurus of plant traits would be  `http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass`.  
<!-- units and factor levels-->
Similarly, standardisation of units can be achieved by relying on SI convention or by relating units to unambigous concepts via URIs provided by ontologies [@steinberg16; @keil18]. 
For categorial, ordinal or binary traits, the factor levels should conform to expected levels as defined in the trait concept, or be unambiguously defined in the metadata of the dataset (see section metadata below). 
The vocabulary complements the user-defined values in dataset-specific units and factor levels with standardised entries (`traitValueStd` and `traitUnitStd`). 
<!-- double record -->
By suggesting for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation which may refer to different scientific community standards  [as practiced in TRY; @kattge11]. This redundancy of data allows for continuity for data providers while also enabling quality checks and comparability for data curators.

@@

Beyond measurement units or higher taxon information, further information might complement the core data which are related to the individual specimen, the reported fact, measurement or sampling event. 
We propose three extensions of the vocabulary that should be used to describe this information (Fig. 2*d*):

- The `Taxon` extension provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.  (https://terminologies.gfbio.org/terms/ets/pages/#extension-taxon)
- The `MeasurementOrFact` extension  provides terms to describe information at the level of single measurements or reported facts, such as the original literature reference for the reported value, the method of measurement or statistical method of aggregation. It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g. variation in measurement method) or aggregated values (e.g. statistical method), as well as the source of reported facts (e.g. literature source or expert reference). (https://terminologies.gfbio.org/terms/ets/pages/#extension-measurement-or-fact)
- The `Occurrence` extension contains vocabulary to describe information on the observation context of individual specimens, such as sex, life stage or age. This also includes the method of sampling and preservation, as well as the date and geographical location, which provide an important resource to analyse trait variation due to differences in space and time. (https://terminologies.gfbio.org/terms/ets/pages/#extension-occurrence)

<!--Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
 If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).  
Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level. This information can then be used for trait inference at higher taxonomic resolution. <!-- using hierarchical probabilistic matrix factorisation [@schrodt15]. -->
These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets, thus avoiding redundancy and duplication of content. A unique identifier links to these other datasheets, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`). 
Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g. a GBIF URI or a stable identifier references the precise specimen from which the measurement was taken [@groom17; @guntsch17].

Identifiers also can be used to provide a structure within a two-dimensional data table, e.g. to link rows of measurements on the same individual (using `occurrenceID`) or observations at the same location (using `locationID`). The  values of multivariate measurements can be labelled using `measurementID`.

@@

Higher-level information about structure, authorship and data ownership and terms of use should be considered when sharing and working with trait datasets.
Our metadata vocabulary (https://ecologicaltraitdata.github.io/ETS/#metadata-vocabulary) provides terms to label these information. 
In the case of primary measurement data, this information applies to the entire trait dataset, and should be stored along with the published data as metadata [e.g. in a separate file, possibly applying the ecological metadata language, EML, @knb11]. 
Whenever data from different sources are compiled into a trait database, these information must be provided at the measurement level to maintain reference to the original data provider and conditions of re-use of these data. This can be achieved by appending the information as columns to the core dataset, or via an unambiguous `datasetID` and a descriptive `datasetName`.

Wherever possible, the field `traitID` should point to a publicly available, unambiguous trait definition in a published ontology (using its URI). 
If such an external reference does not exist, trait data should always be accompanied by a dataset-specific list of trait concepts. 
Such a controlled vocabulary would, in its simplest form, assign trait names with an unambiguous definition of the trait and an expected format of measured values or reported facts (e.g. units or legit factor levels). Ideally, this definition refers to or refines terms from published trait ontologies.
By providing a minimal vocabulary for trait lists (see https://ecologicaltraitdata.github.io/ETS/#terms-for-trait-definitions), we hope to facilitate the unambiguous definition of traits for trait datasets.
This vocabulary might also prove useful for the future publication of trait ontologies. 
<!--See our online reference for recommendations how to prepare a traitlist for  (https://ecologicaltraitdata.github.io/ETS/thesauri.html). We also provide an example thesaurus on Github (http://ecologicaltraitdata.github.io/TraitDataList/), which has been developed to capture traits assessed for multiple animal taxa within the German Biodiversity Exploratories project [@fischer10]. -->
<!--Open access to science data from publicly funded research is a declared goal of international and national research organisations [@emerson15; @allianceofgermanscienceorganisations10; @theroyalsocietysciencepolicycentre12] and increasingly required by journals and funding agencies to warrant reproducibility of science and foster public benefit. 
Also, factual data such as trait measurements, being not a product of a creative act, do not qualify as intellectual property and copyright does not apply [@carroll15; @creativecommons].
However, there is still no broad consensus achieved in the scientific community whether data from public funding must be generally dedicated to the public domain, or whether the investment in production must be acknowledged by attribution or co-authorship. -->
<!-- cP: Maybe we should put in the appendix an example of dataset+metadata? This should probably go in the metadata section 
FDS: to consider!! -->
